\section{Extending ClimateSOM: Detecting Distributional Shifts in Precipitation Co-variability}\label{sec:agu}

In this section, I continue with the \emph{ensemble variability} facet, sharpening it from visual exploration to a specific statistical question.
%
Climate-change impacts are typically assessed through mean-state changes at individual locations, yet the spatial co-variability \emph{between} regions may reorganize fundamentally under warming in ways that marginal trends alone cannot reveal.
%
Detecting such reorganization demands an explicit comparison of the distributions of spatial realizations across scenarios, something neither mean-state analysis nor EOF-based decompositions are designed to directly support.
%
To address this, I extend the ClimateSOM framework with statistical significance testing capabilities, to determine whether the \emph{distribution} of conditioned spatial precipitation patterns has shifted with statistical significance.
%
% Discovering such distributional shifts is difficult with standard analyses and here we demonstrate how ClimateSOM can be used to detect such changes under climate change.
%
I demonstrate this capability on the co-variability between California's Sacramento Bay-Delta Watershed and U.S. West Coast precipitation under LOCA2-downscaled CMIP6 projections, where I find a strengthening departure from historical relationships that I term ``decoherence,'' and the remainder of this section develops the methodology and findings in detail.
%
Although this extension remains a work in progress, I consider it a true litmus test of ClimateSOM's utility. 
%
Rather than a contrived demonstration, it applies the workflow to a climatologically substantive question, whether and how regional precipitation co-variability reorganizes under warming, and stakes its value on whether it can deliver insight that the standard analysis methods of the domain, such as mean-state trend analysis and EOF-based decomposition would struggle. 
%
It is in this confrontation with a real problem, and with the established tools climate scientists already trust, that the workflow's practical worth is genuinely put to the test.

%% Suggested section heads:
\subsection{Introduction}
\label{sec:introduction}
\input{tex/agu/introduction}
\subsection{Data}
\label{sec:data}
\input{tex/agu/data}
\subsection{Methods}
\label{subsec:agu-methods}
\input{tex/agu/methods}
\subsection{Results}
\label{sec:results}
\input{tex/agu/results}
% \subsection{Discussion}
% \label{sec:discussion}
% \input{tex/agu/discussion}
\subsection{Continued Work}
\label{subsec:agu-conclusion}
\input{tex/agu/conclusion}

